The Concurrent Calculi Formalisation Benchmark

### 2024-02-20 short-paper/short-paper.tex
Minor fixes.

Our benchmark problems (detailed in \cref{app:challenges} and on the website) consider
\emph{in isolation} three key aspects that frequently pose challenges
when mechanising concurrency theory: \emph{linearity}, \emph{scope
  extrusion}, and \emph{coinductive reasoning}.  Mechanisations must
often address several of these aspects at the same time; however, we
see the combination of techniques as a next
step%beyond the scope of this challenge
, as discussed further in \cref{sec:going-beyond}.  Our benchmark
challenges are based on process calculi, as these provide a simple but
realistic showcase of common issues encountered during mechanisation
efforts.
% While the three highlighted aspects are not the sole sources of difficulties, they emerge in most mechanisations of concurrency theory.

@@

\section{The Benchmark and its design}\label{sec:design-discussion}
\alb{Next paragraph needs rewriting}
In this section, we outline the factors considered in designing the benchmark challenges.
We begin with some general remarks, then describe the individual design considerations for each challenge problem, and the evaluation criteria for solutions.

Similarly to the authors of the POPLMark challenges, we seek to
answer several questions:
% \alb{Should we stress we are interested in the \textbf{meta-theory} of process calculi, not}
\begin{enumerate}[label=\textbf{(Q\arabic*)},leftmargin=10mm]
\item\label{item:rq1} What is the current state of the art in tools for mechanising the meta-theory of process calculi?
\item\label{item:rq2} Which techniques and best practices can be recommended when starting new mechanisation projects involving process calculi?
\item\label{item:rq3} Which improvements are needed to make mechanisation tools more user-friendly with regards to the issues faced when mechanising process calculi?
\end{enumerate}
%
Note that the challenges revolve around the mechanization of the
\emph{meta-theory} of process calculi, rather than interpreters or
type checkers for the latter.

@@

\paragraph{Extending and combining the current challenges.}
We have aimed to reduce the overlap of the aspects (linearity, scope
extrusion, coinduction) that our challenges involve.  Most concurrency
theory research needs several of those aspects and in fact there is
anedoctical evidence of the difficulties that such a combination may induce
in a mechanization,
e.g.~\cite{DBLP:conf/forte/ZalakainD21}. Furthermore, one can consider
other constructs that we have chosen not to cover for simplicity.  We
suggest three extensions that would be useful for many applications:
\emph{(1)} adding choice constructs to both type and process levels,
\emph{(2)} introducing recursion and recursive types for typed
systems, and \emph{(3)} incorporating channel delegation in calculi
lacking it.

@@

\paragraph{Addressing new calculi and new features.}
% \todo{Mention the higher-order pi calculus? If so, also
%   mention}

Like the POPLMark and POPLMark Reloaded challenges, our % challenge
is
not meant to be comprehensive: Extending it means proposing new
challenges covering different features of process calculi, or
different proof mechanisation features.  Some interesting aspects of
message-passing calculi %to be explored in further challenges could be
include
multiparty session types~\cite{10.1145/2827695} and
choreographies~\cite{Carbone2013}, as their meta-theory includes
aspects -- \eg well-formedness conditions on global types, partiality
of end-point projection function, \etc -- that we do not address
here. Other revolve around higher-order calculi~\cite{Hirsch2022},
conversation types~\cite{DBLP:journals/tcs/CairesV10},
psi-calculi~\cite{lmcs:696}, or encodings between different
calculi~\cite{DBLP:journals/iandc/Gorla10,
  DBLP:conf/ecoop/ScalasDHY17}.\todo{Should we cite other/different
  encoding papers? Which ones are the ``standard'' citations?}

@@

% \todo{This is a list of interesting research problems, but they are
%   not challenges per se. Trim down.}  On the mechanisation side, the
% present challenge proposes mechanising the proofs of theorems.
Finally, another interesting avenue to explore would be to take
advantage of certain proof assistant features: for example, extracting
certified implementations (\cite{Castro-Perez2021}) or exploiting
proof automation, such as the one offered by the \emph{Hammer}
protocol.
% a more specific
% challenge could be designed with the objective of automating aspects
% of the proofs.

Ultimately, the current challenges can be extended in several worthwhile directions, and we look forward to a future when they indeed are.
% We see the current challenge as setting the foundation for those future extensions.
It is our hope and aim that our challenges will move us closer to a future where the key basic proof techniques are as easy to mechanise as they are to write on paper.

>>>>

Our benchmark problems (detailed in \cref{app:challenges} and on our website) consider
\emph{in isolation} three key aspects that frequently pose challenges
when mechanising concurrency theory: \emph{linearity}, \emph{scope
  extrusion}, and \emph{coinductive reasoning}.  Mechanisations must
often address several of these aspects; however, we
see the combination of techniques as a next
step, as discussed in \cref{sec:going-beyond}.
Our benchmark challenges are based on process calculi, as these provide a simple but
realistic showcase of common issues encountered during mechanisation
efforts.
% While the three highlighted aspects are not the sole sources of difficulties, they emerge in most mechanisations of concurrency theory.

@@

\section{The Benchmark and its Design}\label{sec:design-discussion}
\alb{Next paragraph needs rewriting}
In this section, we outline the factors considered in designing the benchmark challenges.
We begin with some general remarks, then describe the individual design considerations for each challenge problem, and the evaluation criteria for solutions.

Similarly to the authors of the POPLMark challenges, we seek to answer several questions:
\begin{enumerate}[label=\textbf{(Q\arabic*)},leftmargin=10mm]
\item\label{item:rq1} What is the current state of the art in tools for mechanising the meta-theory of process calculi?
\item\label{item:rq2} Which techniques and best practices can be recommended when starting new mechanisation projects involving process calculi?
\item\label{item:rq3} Which improvements are needed to make mechanisation tools more user-friendly with regards to the issues faced when mechanising process calculi?
\end{enumerate}
%
\todo{Frederik: I'm not sure this note is necessary}
Note that the challenges revolve around the mechanisation of the
\emph{meta-theory} of process calculi, rather than interpreters or
type checkers for the latter.

@@

\paragraph{Extending and combining the current challenges.}
We have aimed to reduce the overlap of the aspects (linearity, scope
extrusion, coinduction) that our challenges involve.  Most concurrency
theory research needs several of these aspects and in fact there is
anedoctal evidence of the difficulties that such a combination may induce
in a mechanisation,
e.g.~\cite{DBLP:conf/forte/ZalakainD21}. Furthermore, one can consider
other constructs that we have chosen not to cover for simplicity.  We
suggest three extensions that would be useful for many applications:
\emph{(1)} adding choice constructs to both type and process levels,
\emph{(2)} introducing recursion and recursive types for typed
systems, and \emph{(3)} incorporating channel delegation in calculi
lacking it.

@@

\paragraph{Addressing new calculi and new features.}
Like the POPLMark and POPLMark Reloaded challenges, ours
is not meant to be comprehensive: Extending it means proposing new
challenges covering different features of process calculi, or
different proof mechanisation features.  Some interesting aspects of
message-passing calculi include
multiparty session types~\cite{10.1145/2827695} and
choreographies~\cite{Carbone2013}, as their meta-theory includes
aspects -- \eg well-formedness conditions on global types, partiality
of end-point projection function, \etc -- that we do not address
here. Other revolve around higher-order calculi~\cite{Hirsch2022},
conversation types~\cite{DBLP:journals/tcs/CairesV10},
psi-calculi~\cite{lmcs:696}, or encodings between different
calculi~\cite{DBLP:journals/iandc/Gorla10,
  DBLP:conf/ecoop/ScalasDHY17}.
\todo{Should we cite other/different encoding papers? Which ones are the ``standard'' citations?}

@@

Finally, another interesting avenue to explore would be to take
advantage of certain proof assistant features: for example, extracting
certified implementations (as in \cite{Castro-Perez2021}) or exploiting
proof automation, such as the one offered by the \emph{Hammer}
approach.\todo{citation here?}

Ultimately, the current challenges can be extended in several worthwhile directions, and we look forward to a future when they indeed are.
It is our hope and aim that our challenges will move us closer to a future where the key basic proof techniques are as easy to mechanise as they are to write on paper.
